###   Install R Packages
 
```{r Prelims, include=F, echo=FALSE, results='hide', message=FALSE} 
 FN <- "M0A-installPackages"

 # DO NOT USE THIS
 #if(file.exists('/share/R/x86_64-pc-linux-gnu-library/3.0'))
 #.libPaths('/share/R/x86_64-pc-linux-gnu-library/3.0')

  suppressMessages(require(rmarkdown))
  suppressMessages(require(knitr))
  options(stringsAsFactors=F) 

  # Shotcuts for knitting and redering while in R session
  kk <- function(n='') knit2html(paste("Scripts/t", n, sep=''), envir=globalenv(),
        output=paste('Scripts/',FN,".html", sep=''))
  bb <- function(n='') browseURL(paste('Scripts/',FN,".html", sep=''))

  # The usual shorcuts
  zz <- function(n='') source(paste("Scripts/t", n, sep=''))

  if(file.exists('/Users/francois/Dropbox/SingleCell/Jun2017/R/'))
  WRKDIR <- '/Users/francois/Dropbox/SingleCell/Jun2017/R/'  else
  WRKDIR <- '/mnt100/home/Dropbox/SingleCell/Jun2017/R'

  # do once
  #setwd(WRKDIR)

  # file rmarkdown file management options
  #suppressMessages(dir.create('cache/M0A', recursive=T))
  #opts_chunk$set(cache.path='cache/M0A/')
  suppressMessages(dir.create('figure/M0A', recursive=T))
  opts_chunk$set(fig.path='figure/M0A/')

```
<!-- ***************************************************** -->

``` 
[R blog](https://www.r-bloggers.com/how-to-install-r-on-linux-ubuntu-16-04-xenial-xerus/)
 sudo echo "deb http://cran.rstudio.com/bin/linux/ubuntu trusty/" | sudo tee -a /etc/apt/sources.list

 gpg --keyserver keyserver.ubuntu.com --recv-key E084DAB9

 gpg -a --export E084DAB9 | sudo apt-key add -

 sudo apt-get update
 sudo apt-get install r-base r-base-dev

...
```

```{r ldGenomics, echo=T, eval=T}
  #install.packages('S4Vectors')
  # or 
  #R CMD INSTALL BiocGenerics_0.20.0.tar.gz
  #R CMD INSTALL S4Vectors_0.12.1.tar.gz 

 RequiredLibs.vec <- 
  c('Rsamtools', 'GenomeInfoDb', 'GenomicRanges', 'GenomicAlignments',
    "rmarkdown", "knitr", "GenomicRanges", 
    "rtracklayer", # requires libxml2
    "BSgenome.Hsapiens.UCSC.hg19", #  requires libxml2
    "biomaRt", "minfiData", #  requires libxml2
    "AnnotationHub",
    "BSgenome", "yeastRNASeq", 
    "TxDb.Hsapiens.UCSC.hg19.knownGene", "ALL",
    "GEOquery", "airway", "yeastRNASeq", "ShortRead","leeBamViews",
    'ruv', 'sva','QDNAseq','QDNAseq.hg19',
    'dplyr', 'DNAcopy')
    # 'RAmazonS3', 'aws.s3')  ### THESE ODNT WORK!

 InstalledLibs.vec <- installed.packages()[,1]
 ToIntall.vec <- setdiff(RequiredLibs.vec,InstalledLibs.vec)
 
 # Force install of ALL using sudo
 ToIntall.vec <- RequiredLibs.vec

 if(length(ToIntall.vec)){
  source("http://www.bioconductor.org/biocLite.R")
  biocLite(ToIntall.vec)
 }

require(devtools)
install_github("RS3","Gastrograph")

 # GEO
 #biocLite('GEOquery', suppressUpdates=T) 
 #biocLite('hgu133plus2.db', suppressUpdates=T) 
```

```{r}
 ## UP TO HERE
 knit_exit()
```


```{r ldCNA, eval=F}
 # R CMD INSTALL Rhtslib_1.6.0.tar.gz
 # # R CMD INSTALL bamsignals_1.6.0.tar.gz 

  #biocLite('AneuFinder')
  #install.packages('AneuFinder')

  [aneufinder](https://github.com/ataudt/aneufinder)
  library(devtools)
  install_github("ataudt/aneufinderData")

  install_github('ataudt/aneufinder')

  biocLite('QDNAseq')

```{r ldPredictive, eval=F}

 RequiredLibs.vec <- 
  c("gpls", "logicFS", "vbmp",'pROC','pamr','randomForest',
    'kernlab', 'gbm', 'xgboost')

 InstalledLibs.vec <- installed.packages()[,1]
 ToIntall.vec <- setdiff(RequiredLibs.vec,InstalledLibs.vec)

 if(length(ToIntall.vec)){
  source("http://www.bioconductor.org/biocLite.R")
  biocLite(ToIntall.vec)
 }

```


```{r ldOthrt, eval=F}
 install.packages("caret", dependencies = c("Depends", "Suggests"))

 # Do we need to install these?
 biocLite('doMC',  suppressUpdates=T) 

###2: package ‘parallel’ is a base package, and should not be updated 
#biocLite('parallel', suppressUpdates=T) 
biocLite('SDDA',
    lib.loc='/home/ubuntu/Dropbox/R/x86_64-pc-linux-gnu-library/3.3/')


Install.packages('/home/ubuntu/Dropbox/R/SDDA_1.0-5.tar.gz', lib='/home/ubuntu/Dropbox/R/x86_64-pc-linux-gnu-library/3.3')


```
<!-- ***************************************************** -->

<!-- To run
nohup sudo  Rscript -e "knitr::knit2html('M0A-installPackages.Rmd')" > M0A-installPackages.log  &
-->

