###   Install R Packages
 
<style type="text/css"> body, td { font-size: 14px; } code.r{ font-size: 12px; } pre { font-size: 12px } </style>

```{r Prelims, include=F, echo=FALSE, results='hide', message=FALSE} 

 FN <- "M0B-installPackages"

  suppressMessages(require(rmarkdown))
  suppressMessages(require(knitr))
  options(stringsAsFactors=F) 

  # Shotcuts for knitting and redering while in R session
  kk <- function(n='') knit2html(paste("Scripts/t", n, sep=''), envir=globalenv(),
        output=paste('Scripts/',FN,".html", sep=''))
  bb <- function(n='') browseURL(paste('Scripts/',FN,".html", sep=''))

  # The usual shorcuts
  zz <- function(n='') source(paste("Scripts/t", n, sep=''))

  if(file.exists('/Users/francois/Dropbox/SingleCell/Jun2017/R/'))
  WRKDIR <- '/Users/francois/Dropbox/SingleCell/Jun2017/R/'  else
  WRKDIR <- '/mnt100/home/Dropbox/SingleCell/Jun2017/R'

  # do once
  #setwd(WRKDIR)

  # file rmarkdown file management options: cache, figures
  #cache_DIR <- file.path(WRKDIR, 'Scripts', 'cache/M0B/')
  #suppressMessages(dir.create(cache_DIR, recursive=T))
  #opts_chunk$set(cache.path=cache_DIR)

  #figure_DIR <- file.path(WRKDIR, 'Scripts', 'figure/M0B/')
  #suppressMessages(dir.create(figure_DIR, recursive=T))
  #opts_chunk$set(fig.path=figure_DIR)

```
<!-- ***************************************************** -->

### Install additional packages

Install all packages used in
[Hicks](https://github.com/stephaniehicks/scBatchPaper)

#### NOTE: could not install 'mvoutlier', 'destiny' 

Cause 'http://ab-initio.mit.edu/nlopt/nlopt-2.4.2.tar.gz' is non-responsive.
Can download nlopt-2.4.2.tar.gz from 
https://osdn.net/projects/sfnet_daetools/downloads/gnu-linux-libs/nlopt-2.4.2.tar.gz/

To install:
* tar xvfz nlopt-2.4.2.tar.gz
* cd nlopt-2.4.2
* ./configure
* makes
* make install

Then try installing  'mvoutlier' and 'destiny'  again.  **Still Fails!!!**

#### FIXED after ab-initio.mit.edu cma eback online:
* had to "make uninstall" nlopt-2.4.2 first.
* then biocLite(c(mvoutlier', 'destiny'))
* Done!

```{r installPAckages, eval=T}
 RequiredLibs.vec <- unique(
  c('R.utils', 'readxl', 'Rtsne', 'mvoutlier', 'destiny', 'scran', 'BiocParallel',
    'Biobase', 'GenomicRanges', 'vcd', 'stringr', 'reshape2',
    'tidyr', 'plyr', 'dplyr', 'irlba', 'readr', 'scater',
    'knitr', 'ggplot2', 'grid', 'cowplot', 'Biobase', 'GenomicRanges', 
    'stringr', 'reshape2', 'tidyr', 'RColorBrewer', 'plyr', 'dplyr', 
    'irlba', 'rafalib' , 'rafalib', 'SummarizedExperiment', 'MultiAssayExperiment', 
    'scRNASeqMouseShalekBMDCs', 'scRNASeqHCT116CellLinesWu', 'scRNASeqMouseShalekBMDCs', 
    'scRNASeqHCT116CellLinesWu', 'scRNASeqMouseShalekBMDCs', 'scRNASeqHCT116CellLinesWu', 
    'scRNASeqMouseShalekBMDCs', 'scRNASeqHCT116CellLinesWu', 'scRNASeqMouseShalekBMDCs', 
    'scRNASeqHCT116CellLinesWu', 'scRNASeqMouseDengMonoAllelic', 'scRNASeqMouseJaitinSpleen',
    'scRNASeqMouseKumarPSC', 'SummarizedExperiment', 'scRNASeqHumanPatelGlioblastoma', 
    'scRNASeqMouseShalekDendritic', 'scRNASeqMouseTreutleinLineage', 'scRNASeqHumanBosePrinting', 
    'scRNASeqMouseBurnsInnerEar', 'scRNASeqHumanGuoGermCells', 'scRNASeqMouseKowalczykAging', 
    'scRNASeqHumanLengOscillatoryGenes', 'scRNASeqDaniSajitaSeurat', 'scRNASeqMouseZeiselCortex', 
    'scRNASeqHumanPatelGlioblastoma'))


 InstalledLibs.vec <- installed.packages()[,1]
 ToIntall.vec <- setdiff(RequiredLibs.vec,InstalledLibs.vec)


 #ToIntall.vec <- c('nloptr', 'lme4', 'pbkrtest', 'car', 'VIM', 'robCompositions')
 

 if(length(ToIntall.vec)){
  source("http://www.bioconductor.org/biocLite.R")
  biocLite(ToIntall.vec)
 }

```

```{r installGithub, eval=F}
 require(devtools)
 install_github("willtownes/patel2014gliohuman")
 install_github("stephaniehicks/trapnell2014myoblasthuman")

```

### Parameter settings:
  * WRKDIR = `r WRKDIR`
  * FN = `r FN`
  * Scripts = Scripts
  * RUN DATE = `r date()`

```{r, echo=F}
 sessionInfo()
### STOP HERE
  knit_exit()
```


<!-- To run

nohup sudo Rscript -e "knitr::knit2html('M0B-installPackages.Rmd')" > M0B-installPackages.log  &

-->


